\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}

\newcommand{\St}{S_T}
\newcommand{\Szero}{S_0}
\newcommand{\Fzero}{F_0}
\newcommand{\Ft}{F_t}
\newcommand{\K}{K}
\newcommand{\pnl}{\mathrm{P\&L}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\E}{\mathbb{E}}

\newenvironment{solution}{\par\noindent\textbf{Solution.}\ }{\par}

\title{S181M116 Derivatives and Alternative Investments\\
Practice Assignment 1 with Solutions}
\author{Based on Lecture 1 and Lecture 2, Sections 1--3}
\date{}

\begin{document}
\maketitle

\textbf{Instructions.} These questions move from basic definitions to integrated
hedging and pricing problems. Unless otherwise stated, ignore transaction costs,
taxes, bid-ask spreads, and default risk. Use continuous compounding when the
question gives an exponential formula and simple compounding when the question
explicitly says ``simple annual rate.''

\section*{Questions}

\subsection*{Question 1 (Easy): Identifying derivatives}
For each item, say whether it is a derivative. If it is, identify the underlying
variable.
\begin{enumerate}
    \item A contract that pays the holder $\max(S_T-100,0)$ in three months.
    \item A share of common stock.
    \item A contract to exchange EUR 1 million for USD at a fixed exchange rate
    in six months.
    \item A corporate bond.
    \item A contract that pays when average temperature in July exceeds a fixed
    threshold.
\end{enumerate}

\subsection*{Question 2 (Easy): Payoff, profit, and value}
A student buys a European call option with strike $K=50$ for a premium of $4$.
At maturity, the stock price is $S_T=55$.
\begin{enumerate}
    \item What is the call payoff?
    \item Ignoring interest on the premium, what is the buyer's profit?
    \item Explain in one sentence why payoff and profit are not the same.
\end{enumerate}

\subsection*{Question 3 (Easy): Market structure}
Give two advantages and two disadvantages of exchange-traded derivatives
relative to OTC derivatives. Your answer should mention standardization,
clearing, liquidity, and customization.

\subsection*{Question 4 (Easy): Forward payoffs}
A forward contract has delivery price $K=50$. Compute the long and short forward
payoffs at maturity when $S_T=42$, $S_T=50$, and $S_T=60$.

\subsection*{Question 5 (Easy): Choosing the hedge direction}
For each exposure, say whether the firm should use a short hedge or a long hedge.
Briefly explain why.
\begin{enumerate}
    \item A farmer will sell wheat after harvest.
    \item An airline will buy jet fuel in three months.
    \item A U.S. importer must pay EUR 2 million in six months.
    \item An oil producer will sell crude oil next quarter.
\end{enumerate}

\subsection*{Question 6 (Medium): Currency forward hedge}
A U.S. firm must pay GBP 2 million in six months. It enters a long GBP forward
at USD 1.2400 per GBP.
\begin{enumerate}
    \item What dollar payment has the firm locked in?
    \item If the spot exchange rate at maturity is USD 1.3100 per GBP, what is
    the payoff on the forward position?
    \item What is the combined dollar cost of buying GBP in the spot market and
    receiving the forward payoff?
\end{enumerate}

\subsection*{Question 7 (Medium): No-arbitrage forward price}
A non-dividend-paying asset has $S_0=75$. The simple annual risk-free rate is
$4\%$ and maturity is six months.
\begin{enumerate}
    \item Compute the no-arbitrage forward price using simple compounding.
    \item If the quoted forward price is $78.20$, describe the arbitrage trade.
    \item What is the locked-in profit per unit of the asset?
\end{enumerate}

\subsection*{Question 8 (Medium): Daily futures settlement and margin}
A trader is long four gold futures contracts. Each contract is for 100 ounces.
Yesterday's settlement price was USD 2,050 per ounce and today's settlement
price is USD 2,038 per ounce. The trader's initial margin account balance was
USD 20,000 and the maintenance margin is USD 16,000.
\begin{enumerate}
    \item Compute today's futures P\&L.
    \item What is the margin account balance after daily settlement?
    \item Is there a margin call? If so, how much must be deposited to restore
    the account to the initial margin level?
\end{enumerate}

\subsection*{Question 9 (Medium): Option payoffs and profits}
A call and a put have the same strike, $K=100$. The call premium is $6$ and the
put premium is $4$. Fill in the payoff and profit for the buyer of each option
when $S_T=92$, $S_T=100$, and $S_T=112$.

\subsection*{Question 10 (Medium): Hedging, speculation, arbitrage, or convergence trading}
Classify each trade as hedging, speculation, true arbitrage, or convergence
trading. Briefly justify your answer.
\begin{enumerate}
    \item A farmer sells corn futures because she will sell corn after harvest.
    \item A trader buys crude oil futures because he expects oil prices to rise.
    \item A trader borrows cash, buys an asset, and shorts an overpriced forward
    to lock in a riskless profit under the assumptions of the no-arbitrage model.
    \item A fund buys a cheap bond and shorts a similar expensive bond, expecting
    their prices to converge, but the trade can lose money before convergence.
\end{enumerate}

\subsection*{Question 11 (Hard): Effective sale price in a short hedge}
A firm will sell a commodity at date 2. At date 1, it shorts futures at
$F_1=6.20$. At date 2, the spot price is $S_2=5.85$ and the futures price is
$F_2=5.95$.
\begin{enumerate}
    \item Compute the gain or loss on the short futures position.
    \item Compute the effective sale price.
    \item Compute the basis at date 2 and verify that the effective sale price is
    $F_1+b_2$.
    \item Explain why this hedge is not the same as selling the physical asset at
    $F_1$ in advance.
\end{enumerate}

\subsection*{Question 12 (Hard): Effective purchase price in a long hedge}
A manufacturer will buy copper at date 2. At date 1, it goes long futures at
$F_1=3.70$. At date 2, the spot price is $S_2=3.92$ and the futures price is
$F_2=3.88$.
\begin{enumerate}
    \item Compute the gain or loss on the long futures position.
    \item Compute the effective purchase price.
    \item Compute the basis at date 2 and verify that the effective purchase
    price is $F_1+b_2$.
    \item If the basis at date 2 had been negative, would the effective purchase
    price have been above or below $F_1$?
\end{enumerate}

\subsection*{Question 13 (Hard): Cross hedge and minimum-variance hedge ratio}
An airline will buy 2,100,000 gallons of jet fuel. It cross hedges with
heating-oil futures because the heating-oil contract is more liquid. Each
futures contract covers 42,000 gallons. Historical data give
\[
    \rho=0.75,\qquad \sigma_S=0.20,\qquad \sigma_F=0.16,
\]
where $\Delta S$ is the change in the jet-fuel price and $\Delta F$ is the
change in the futures price.
\begin{enumerate}
    \item Compute the minimum-variance hedge ratio.
    \item How many futures contracts should the airline use?
    \item Should the airline go long or short futures?
    \item Name two sources of residual risk that remain after the hedge.
\end{enumerate}

\subsection*{Question 14 (Hard): Forward pricing with income and contract value}
\begin{enumerate}
    \item A stock has $S_0=100$, the present value of known dividends during the
    life of the forward is $I=2.50$, the continuously compounded risk-free rate
    is $5\%$, and $T=1$. Compute the no-arbitrage forward price.
    \item A stock index has $S_0=100$, continuous dividend yield $q=2\%$,
    continuously compounded risk-free rate $r=5\%$, and $T=1$. Compute the
    no-arbitrage futures/forward price.
    \item You are long an old forward contract with delivery price $K=48$.
    A new forward with the same maturity has delivery price $F_0=52$.
    The continuously compounded risk-free rate is $3\%$ and remaining maturity
    is six months. Compute the value of the old long forward.
\end{enumerate}

\subsection*{Question 15 (Hard): Commodity pricing, convenience yield, and expected spot prices}
A consumption commodity has spot price $S_0=80$. The continuously compounded
risk-free rate is $5\%$, proportional storage cost is $1\%$, convenience yield
is $4\%$, and maturity is six months.
\begin{enumerate}
    \item Compute the futures price implied by
    $F_0=S_0e^{(r+u-y)T}$.
    \item Compute the futures price if the convenience yield were zero.
    \item Explain why a futures price below the no-convenience-yield cost of
    carry is not automatically an arbitrage opportunity for a consumption
    commodity.
    \item Separately, suppose $\E(S_T)=84$ and the required expected return on
    the commodity is $k=7\%$. Using $F_0=\E(S_T)e^{(r-k)T}$, compute the futures
    price implied by the risk-premium relation. Is it above or below the
    expected future spot price?
\end{enumerate}

\newpage
\section*{Solutions}

\subsection*{Solution 1}
\begin{solution}
\begin{enumerate}
    \item Derivative. The underlying is the stock price $S_T$.
    \item Not a derivative. It is the underlying asset itself.
    \item Derivative. The underlying is the EUR/USD exchange rate.
    \item Not usually treated as a derivative in this course. It is a debt
    security, though its value depends on interest rates and default risk.
    \item Derivative. The underlying is the July temperature index or average
    temperature measure specified in the contract.
\end{enumerate}
\end{solution}

\subsection*{Solution 2}
\begin{solution}
The call payoff is
\[
    \max(S_T-K,0)=\max(55-50,0)=5.
\]
The buyer's profit is payoff minus premium:
\[
    5-4=1.
\]
Payoff and profit differ because the payoff is only the terminal cash flow,
whereas profit also subtracts the initial cost and any financing cost.
\end{solution}

\subsection*{Solution 3}
\begin{solution}
Exchange-traded derivatives are standardized and usually liquid. Standardization
makes them easier to trade anonymously, and the clearing house reduces bilateral
counterparty risk through margin, daily settlement, and default resources.

The main disadvantages are lack of customization and possible basis mismatch.
The exchange contract may not have the exact maturity, quantity, location,
currency, or payoff that the hedger needs. OTC contracts can be tailored more
closely, but they bring greater valuation, legal, liquidity, and counterparty
risk complications.
\end{solution}

\subsection*{Solution 4}
\begin{solution}
The long forward payoff is $S_T-K$ and the short forward payoff is $K-S_T$.

\begin{center}
\begin{tabular}{@{}ccc@{}}
\toprule
$S_T$ & Long payoff $S_T-K$ & Short payoff $K-S_T$ \\
\midrule
42 & $42-50=-8$ & $50-42=8$ \\
50 & $50-50=0$ & $50-50=0$ \\
60 & $60-50=10$ & $50-60=-10$ \\
\bottomrule
\end{tabular}
\end{center}
\end{solution}

\subsection*{Solution 5}
\begin{solution}
\begin{enumerate}
    \item Short hedge. The farmer will sell wheat later and is hurt by falling
    wheat prices.
    \item Long hedge. The airline will buy fuel later and is hurt by rising fuel
    prices.
    \item Long hedge in EUR. The importer must buy EUR later and is hurt if EUR
    appreciates against USD.
    \item Short hedge. The oil producer will sell crude oil later and is hurt by
    falling oil prices.
\end{enumerate}
\end{solution}

\subsection*{Solution 6}
\begin{solution}
The locked-in dollar payment is
\[
    2{,}000{,}000\times 1.2400=\$2{,}480{,}000.
\]
If the terminal spot rate is 1.3100, the payoff on the long GBP forward is
\[
    (1.3100-1.2400)\times 2{,}000{,}000=\$140{,}000.
\]
Buying GBP in the spot market costs
\[
    2{,}000{,}000\times 1.3100=\$2{,}620{,}000.
\]
After the forward payoff, the net dollar cost is
\[
    2{,}620{,}000-140{,}000=\$2{,}480{,}000.
\]
\end{solution}

\subsection*{Solution 7}
\begin{solution}
With simple compounding,
\[
    F_0=S_0(1+rT)=75(1+0.04\times 0.5)=75(1.02)=76.50.
\]
The quoted forward price 78.20 is too high relative to no-arbitrage value. The
cash-and-carry arbitrage is:
\begin{enumerate}
    \item Borrow 75.
    \item Buy the asset.
    \item Short the forward at 78.20.
    \item At maturity, deliver the asset into the forward, receive 78.20, and
    repay $75(1.02)=76.50$.
\end{enumerate}
The locked-in profit per unit is
\[
    78.20-76.50=1.70.
\]
\end{solution}

\subsection*{Solution 8}
\begin{solution}
The total position is
\[
    NQ=4\times 100=400
\]
ounces. For a long futures position,
\[
    \pnl=NQ(F_t-F_{t-1})=400(2038-2050)=400(-12)=-\$4{,}800.
\]
The margin account balance after settlement is
\[
    20{,}000-4{,}800=15{,}200.
\]
Since 15,200 is below the maintenance margin of 16,000, there is a margin call.
To restore the account to the initial margin level of 20,000, the trader must
deposit
\[
    20{,}000-15{,}200=\$4{,}800.
\]
\end{solution}

\subsection*{Solution 9}
\begin{solution}
For the call, payoff is $\max(S_T-100,0)$ and profit is payoff minus 6. For the
put, payoff is $\max(100-S_T,0)$ and profit is payoff minus 4.

\begin{center}
\begin{tabular}{@{}ccccc@{}}
\toprule
$S_T$ & Call payoff & Call profit & Put payoff & Put profit \\
\midrule
92  & 0  & -6 & 8 & 4 \\
100 & 0  & -6 & 0 & -4 \\
112 & 12 & 6  & 0 & -4 \\
\bottomrule
\end{tabular}
\end{center}

The table also illustrates why option payoff and option profit are different:
the buyer must first recover the premium.
\end{solution}

\subsection*{Solution 10}
\begin{solution}
\begin{enumerate}
    \item Hedging. The farmer has an existing business exposure to a future sale
    price and uses futures to reduce that risk.
    \item Speculation. The trader is taking a directional view on crude oil
    prices without an offsetting underlying business exposure.
    \item True arbitrage under the model assumptions. The trade locks in a
    positive profit with no price risk if the no-arbitrage assumptions hold.
    \item Convergence trading. The trade depends on prices converging and can
    lose money before convergence, so it is not true arbitrage.
\end{enumerate}
\end{solution}

\subsection*{Solution 11}
\begin{solution}
The short futures gain is
\[
    F_1-F_2=6.20-5.95=0.25.
\]
The effective sale price is the physical sale price plus the futures gain:
\[
    S_2+(F_1-F_2)=5.85+0.25=6.10.
\]
The date-2 basis is
\[
    b_2=S_2-F_2=5.85-5.95=-0.10.
\]
Therefore
\[
    F_1+b_2=6.20-0.10=6.10,
\]
which matches the effective sale price.

This hedge is not the same as pre-selling the physical commodity at $F_1$.
The firm still sells the physical commodity later at the spot price $S_2$. The
futures position adds a gain or loss. If $b_2$ is uncertain, the effective price
is uncertain even though the futures position reduces price risk.
\end{solution}

\subsection*{Solution 12}
\begin{solution}
The long futures gain is
\[
    F_2-F_1=3.88-3.70=0.18.
\]
The effective purchase price is the physical purchase price minus the futures
gain:
\[
    S_2-(F_2-F_1)=3.92-0.18=3.74.
\]
The date-2 basis is
\[
    b_2=S_2-F_2=3.92-3.88=0.04.
\]
Therefore
\[
    F_1+b_2=3.70+0.04=3.74,
\]
which matches the effective purchase price.

If $b_2$ had been negative, the effective purchase price would have been below
$F_1$. For a long hedge, positive basis raises the effective purchase price
above the initial futures price, while negative basis lowers it.
\end{solution}

\subsection*{Solution 13}
\begin{solution}
The minimum-variance hedge ratio is
\[
    h^*=\rho\frac{\sigma_S}{\sigma_F}
    =0.75\frac{0.20}{0.16}
    =0.9375.
\]
The number of contracts is
\[
    N^*=h^*\frac{Q_A}{Q_F}
    =0.9375\frac{2{,}100{,}000}{42{,}000}
    =0.9375(50)
    =46.875.
\]
The airline should use about 47 contracts. Since the airline will buy fuel later
and is worried about rising prices, it should go long futures.

Residual risk remains because jet fuel and heating oil are not the same asset,
so their prices may not move one-for-one. There can also be timing mismatch,
location mismatch, changing correlations, contract-size rounding, and basis risk
when the hedge is closed.
\end{solution}

\subsection*{Solution 14}
\begin{solution}
For known cash income,
\[
    F_0=(S_0-I)e^{rT}=(100-2.50)e^{0.05}
    =97.50e^{0.05}\approx 102.50.
\]
For a known continuous yield,
\[
    F_0=S_0e^{(r-q)T}=100e^{(0.05-0.02)(1)}
    =100e^{0.03}\approx 103.05.
\]
For the old long forward,
\[
    f=(F_0-K)e^{-rT}
    =(52-48)e^{-0.03\times 0.5}
    =4e^{-0.015}
    \approx 3.94.
\]
The old long forward is valuable because it lets the holder buy at 48 when a new
forward would lock in a purchase price of 52.
\end{solution}

\subsection*{Solution 15}
\begin{solution}
With storage cost and convenience yield,
\[
    F_0=S_0e^{(r+u-y)T}
    =80e^{(0.05+0.01-0.04)(0.5)}
    =80e^{0.01}
    \approx 80.80.
\]
If the convenience yield were zero,
\[
    F_0=80e^{(0.05+0.01)(0.5)}
    =80e^{0.03}
    \approx 82.44.
\]
A futures price below the no-convenience-yield cost of carry is not automatically
an arbitrage opportunity for a consumption commodity because physical ownership
may provide convenience benefits. Also, reverse cash-and-carry can be difficult
when the commodity is hard to short, hard to borrow, costly to store, or needed
for production.

Using the risk-premium relation,
\[
    F_0=\E(S_T)e^{(r-k)T}
    =84e^{(0.05-0.07)(0.5)}
    =84e^{-0.01}
    \approx 83.16.
\]
This futures price is below the expected future spot price because the commodity
has positive systematic risk, so $k>r$.
\end{solution}

\end{document}
